On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-06-01 Manuscript/manuscript.tex
removed resolved

Reaction time in sprint and hurdle events has been studied by many authors.
\citet{pain2007sprint} examined
reaction times for nine male athletes.  The
fastest athlete of the nine had, on average (even with their two fastest times 
removed), a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. 
In a study commissioned by International Association of Athletics Federations 
(IAAF), \citet{komi2009iaaf} recommended that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
\eds{are hurdlers generally considered a subset of sprinters or are they 
considered a separate type of athlete?  Specifically did the study above 
recommend the barrier be lowered for both sprinters and hurdlers?
You also refer to sprinters elsewhere in the manuscript.  Are those instances
meant to refer to both sprinters and hurdlers, as well?}
\of{Yes I would say that hurdlers are a subset of sprinter.  The reaction time
barrier for the men's 100 meters and men's 110 meter hurdle are both 0.1 seconds.
The paper recommended for the reaction time barrier to be decreased to roughly 0.8
seconds.  It does not reference any distinction between hurdlers/sprinters, so I
think they can be grouped together.  In my personal opinion I don't see why the
reaction time for hurdlers versus 100m dash would differ significantly.  In terms
of what I wrote in the paper, I want to stay accurate to what is in the study
but also generalize studies involving reaction times to include hurdlers.  I believe
many of the studies examining the reaction time barrier from a physiological
perspective to find the minimum possible reaction time do not deal with hurdles.
Thus, I don't know if I can change what I wrote to be both accurate and understandable,
but what do you both think?}
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body
movement. This could provide a more accurate representation of reaction time, 
and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been proved to be a reliable
measure of reaction time \citep{mudric2015evaluation}.

@@

There are two types of data that are explored throughout this paper: data from
1999-2022 for every world championship, and data from athletes who competed in
the 2022 World Championships and another competition (either 2019 World 
Championships or a 2022 national competition typically held within three months 
of the 2022 World Championships). 
\eds{is this true? are you still using the 2019 world championships data for the
rank-based analyses, or only the national competitions?}
\of{Yes the ranked based comparison has 2 subsections: 2022 national vs 2022
international and 2019 international vs 2022 international.  Should I instead say
that there are three types of data across 2 sections or something like that?}
\eds{I think this is fine here, but it is not at all clear that there are two
subsections in the rank-based results for 2022 national vs 2022
international and 2019 international vs 2022 international.}
\of{Currently under the Methods/Results section I have GLMM and rank based comparison
do you think I need subsections under rank based comparison, or is that too much
compartmentalization?  Or should I instead say there are 3 types of data that are
used in two different methods, or something like that?}
The first data set is used for the GLMM analyses described in 
Section~\ref{sec:glmm}, while the second collection of data is used for the 
rank-based comparisons described in Section~\ref{sec:rank}.

@@

Data was taken from the World Athletics 
%\eds{is there a website you can cite here?}
and covers the men's 110 meter hurdles from 1999 to 2022; the source can be 
found in the Appendix.
Although women's data is available, we chose not to include women's data in the
analyses to follow due to the numerous papers citing differences in men and
women's reaction times \citep[e.g.,][]{lipps2011implications, 
babicc2009reaction, panoutsakopoulos2020gender}.
We focus on the reaction times during the semifinal and final heats only, as 
reaction times from preliminary heats are often not as fast as those in later 
heats \citep[e.g.,][]{zhang2021correlation}.

@@

\eds{I wonder if this paragraph would be better placed in the introduction?}
\of{I think it could fit in the brief section between the start of the Data section
and the GLMM subsection in the paragraph starting with "There are two types..."}
\eds{With the following newly added sentences, I think we can leave this paragraph here.}
\of{I agree I think this works well.}
We can also see from Figure~\ref{fig:Boxplots} that the reaction times clearly
differ from year to year. The venue of the World Championships changes each 
year, so weather or climate related factors such as humidity, precipitation, 
elevation, may indeed play a role, but additionally, both the technology and 
false start penalization rules have changed over the study period.

@@

\eds{this is where you can put the relevant 
information about the other competitions that was commented out in the 
introduction.}
\eds{Can you show some descriptive statistics for this data?  Total number of 
clusters? Average number of international vs national competitions across 
athletes/clusters?}
\of{Is the one below what you're talking about?}
\eds{Yes, this is the part from the intro I was referring to, but you need to 
edit this paragraph now so it all works together.}

The dataset of the four United States athletes 
is too small to perform any meaningful statistical analysis, but
the data was expanded to include reaction times for athletes who competed at 
national competitions (many of which were held from May-July of 2022). Once the 
data was collected, each athlete was designated as a cluster and the stage of
competition (national or international) was specified for each competition 
within the cluster (athlete). First we can compare the reaction times of athletes who
competed at the World Track and Field Championships in both 2019 and 2022, and 
then we can compare how athletes performed at the 2022 championships relative 
to their national/qualifiers meet. In order to reduce any chances
of athletes improving reaction time, we can also look at data from qualifying meets.
Almost every athlete had to qualify by proving they are elite at the national level, 
usually by coming in at least fourth. Thus, data was compiled on athletes who competed
at the 2022 World Track and Field Championships on what these athletes also ran 
at their country's national championship. If there was any argument that athletes
have improved their reaction time from 2019 to 2022, it should now be considered
nearly null, as the national competitions took place between June and July 2022 
and the World Championships were in August. The equipment used to measure reaction
time cannot be guaranteed to be the same from national to international
competition, but the time span between meets has been reduced considerably.

@@

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the final model we chose was a gamma mixed-effect 
model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. A random effect for the venue or year was also found to be necessary, 
which makes intuitive sense given the discussion in Section~\ref{sec:dataworld}. 
Let $Y_{ij}$ be the reaction time of observation~$j$ in year~$i$.
\eds{we know that many of the athletes compete in multiple world championships,
  how is this athlete clustering being accounted for here?}
\jy{We could add an athlete random effect too, if the players are identified in
  the data.}
\jy{Owen, do we have athlete id in the data?}
Conditional on a venue effect $z_i$ of year~i,  $Y_{ij}$ is modeled by 
gamma distribution. In the classic generalized linear model (GLM) sense,
$Y_{ij}$ has conditional mean
\[
g\{E(Y_{ij} | z_i)\} = \alpha + z_i,
\]
and dispersion parameter $\phi$, where $g$ is a known link function and
$z_i$ is a normally distributed random effect with mean zero and
variance~$\sigma^2$. Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time data, the logarithm 
link function worked better for our data and additionally ensures the positivity
of the conditional mean. The gamma mixed-effect model can be fit with the 
\texttt{glmer()} function in R package \texttt{lme4} \citep{lme4}.

@@

\of{Add something like this to the start of the paragraph?}
We may want to consider using all four data sets for this problem because
comparing all the data to the data excluding 2022 can give us an
assessment of how impactful 2022 was. We hope to see that by including 2022,
the probability of observing a significantly low reaction time increases. 
\eds{Agree with JY, but you need to describe what these four datasets are.}
\of{I added a description of the four datasets in the Data paragraph right below
where it says <Thus, we have four datasets>}
Table~\ref{tab:Gamma_parameters} summarizes the results from the gamma model 
from different data sets and shows the standard error of the year
effect increases for both the men's final and men's pooled data sets when
including 2022.  The men's final data venue effect error increased from 0.439
to 0.504 and from 0.0495 to 0.0554 for the men's pooled data.  This demonstrates
that the inclusion of 2022 increased the randomness of the data and motivates
the rank-based study (see Section~\ref{subsec:Results_Rank}).

@@

Using the gamma function with link set to log, the probability of the reaction 
time being less than 0.1 seconds can be calculated to see if Allen's 
disqualification was an anomaly. 
\eds{explain how simulation was used here.} \of{The methods section already
details how the calculations are performed, can you please elaborate what you
are exactly looking for.}
Calculations were performed in R using the methods described in 
section~\ref{sec:glmm} with the use of the \citet{lme4} package.

At a size of $n=10,000,000$, \eds{use different letter as $n$ was already used 
for number of athletes, $B?$} \of{Could I just say "At a size of 10,000,000"? 
Otherwise I am fine with using "B" instead of "n"}

@@

%Citation needed in this section
Since 1985 Seiko Holding Corporations has served every World Athletics 
Championships as the official timer \citep{wa2022seiko}.  Since 1985, the 
technology and the ability
to accurately predict measurements: long jump distances, false starts, total 
time, reaction time, etc. have all dramatically improved.  Seiko did not start 
tracking reaction time as an official measurement until 1999 in the men's and 
women's 100 meter \eds{110 for men?} hurdles. 
Seiko regularly updates their equipment so that they provide cutting edge
technology to the World Track and Field Championships, the highest stakes in the 
world of running outside of the Olympics. Seiko's technology for detecting 
reaction times relies on the pressure that athletes exert on the plate when they 
push off.  Their systems measure the time differential
between when the ``gun" goes off to start the race and when the pressure changes 
\citep{wa2022seiko}.  If an athlete has a reaction time under 0.1 seconds, it 
is deemed a 
false start as it is considered that no athlete can react so quickly 
\citep{Seiko-Timing}.  Thus, a reaction time of for example 0.05, suggests that 
the athlete predicted the gun and it was luck that caused their abnormally high 
time.  At the World Championships level, World Athletics and Seiko want to remove 
that element of luck and thus impose the 0.1 second barrier. In 2013, Seiko 
upgraded their timing equipment for sprinting events (110 meter hurdles falls 
under this category) \citep{wa2013backtage}. Since 1999, the
slowest \eds{highest as in fastest or slowest? Neither seems consistent with 
the boxplots in Figure 1} \of{Resolved} reaction 
times in the men's 110-meter hurdle were recorded in 2013.  That is not to say
 that the higher 
times were caused by Seiko's equipment, but the two may be related.  It is worth 
noting that prior
to 2022, Seiko again upgraded its technology but for its jump management system.
\eds{I would argue that this paragraph should go in the discussion as there are 
no results presented here}

>>>>

Reaction time in sprint and hurdle events has been studied by many authors.
\citet{pain2007sprint} examined
reaction times for nine male athletes.  The
fastest athlete of the nine had, on average (even with their two fastest times 
removed), a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. 
In a study commissioned by International Association of Athletics Federations 
(IAAF), \citet{komi2009iaaf} recommended that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body
movement. This could provide a more accurate representation of reaction time, 
and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been proved to be a reliable
measure of reaction time \citep{mudric2015evaluation}.

@@

There are two types of data that are explored throughout this paper: data from
1999-2022 for every world championship, and data from athletes who competed in
the 2022 World Championships and another competition (either 2019 World 
Championships or a 2022 national competition typically held within three months 
of the 2022 World Championships). 
\eds{is this true? are you still using the 2019 world championships data for the
rank-based analyses, or only the national competitions?}
\of{Yes the ranked based comparison has 2 subsections: 2022 national vs 2022
international and 2019 international vs 2022 international.  Should I instead say
that there are three types of data across 2 sections or something like that?}
\eds{I think this is fine here, but it is not at all clear that there are two
subsections in the rank-based results for 2022 national vs 2022
international and 2019 international vs 2022 international.}
\of{Currently under the Methods/Results section I have GLMM and rank based comparison
do you think I need subsections under rank based comparison, or is that too much
compartmentalization?  Or should I instead say there are 3 types of data that are
used in two different methods, or something like that?} \eds{I prefer
subsections within rank-based comparison.}
The first data set is used for the GLMM analyses described in 
Section~\ref{sec:glmm}, while the second collection of data is used for the 
rank-based comparisons described in Section~\ref{sec:rank}.

@@

Data was taken from the World Athletics 
and covers the men's 110 meter hurdles from 1999 to 2022; the source can be 
found in the Appendix.
Although women's data is available, we chose not to include women's data in the
analyses to follow due to the numerous papers citing differences in men and
women's reaction times \citep[e.g.,][]{lipps2011implications, 
babicc2009reaction, panoutsakopoulos2020gender}.
We focus on the reaction times during the semifinal and final heats only, as 
reaction times from preliminary heats are often not as fast as those in later 
heats \citep[e.g.,][]{zhang2021correlation}.

@@

We can also see from Figure~\ref{fig:Boxplots} that the reaction times clearly
differ from year to year. The venue of the World Championships changes each 
year, so weather or climate related factors such as humidity, precipitation, 
elevation, may indeed play a role, but additionally, both the technology and 
false start penalization rules have changed over the study period.

@@

\eds{Can you show some descriptive statistics for this data?  Total number of 
clusters? Average number of international vs national competitions across 
athletes/clusters?}

\eds{The following paragraph still needs to be smoothed (the paragraph that was 
here and the paragraph that was in the intro still need to be integrated into 
one (or two) cohesive paragraph(s) that accurately describe the two types of 
data that will be used for two rank-based analyses.}
The dataset of the four United States athletes 
is too small to perform any meaningful statistical analysis, but
the data was expanded to include reaction times for athletes who competed at 
national competitions (many of which were held from May-July of 2022). Once the 
data was collected, each athlete was designated as a cluster and the stage of
competition (national or international) was specified for each competition 
within the cluster (athlete). First we can compare the reaction times of 
athletes who
competed at the World Track and Field Championships in both 2019 and 2022, and 
then we can compare how athletes performed at the 2022 championships relative 
to their national/qualifiers meet. In order to reduce any chances
of athletes improving reaction time, we can also look at data from qualifying 
meets. Almost every athlete had to qualify by proving they are elite at the 
national level, usually by coming in at least fourth. Thus, data was compiled on 
athletes who competed at the 2022 World Track and Field Championships on what 
these athletes also ran at their country's national championship. If there was 
any argument that athletes
have improved their reaction time from 2019 to 2022, it should now be considered
nearly null, as the national competitions took place between June and July 2022 
and the World Championships were in August. The equipment used to measure 
reaction time cannot be guaranteed to be the same from national to international
competition, but the time span between meets has been reduced considerably.

@@

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the final model we chose was a gamma mixed-effect 
model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. A random effect for the venue or year was also found to be necessary, 
which makes intuitive sense given the discussion in Section~\ref{sec:dataworld}. 
Let $Y_{ij}$ be the reaction time of observation~$j$ in year~$i$.
%\eds{we know that many of the athletes compete in multiple world championships,
%  how is this athlete clustering being accounted for here?}
%\jy{We could add an athlete random effect too, if the players are identified in
%  the data.}
%\jy{Owen, do we have athlete id in the data?}
Conditional on a venue effect $z_i$ of year~i,  $Y_{ij}$ is modeled by 
gamma distribution. In the classic generalized linear model (GLM) sense,
$Y_{ij}$ has conditional mean
\[
g\{E(Y_{ij} | z_i)\} = \alpha + z_i,
\]
and dispersion parameter $\phi$, where $g$ is a known link function and
$z_i$ is a normally distributed random effect with mean zero and
variance~$\sigma^2$. Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time data, the logarithm 
link function worked better for our data and additionally ensures the positivity
of the conditional mean. The gamma mixed-effect model can be fit with the 
\texttt{glmer()} function in R package \texttt{lme4} \citep{lme4}.

@@

\of{Add something like this to the start of the paragraph?}
We may want to consider analyzing all four data sets separately because
comparing the model based on all the data to the model based on data excluding 
2022 can provide an indication about the impact of the 2022 data.
We hope to see that by including 2022 data,
the probability of observing a significantly low reaction time increases. 
Table~\ref{tab:Gamma_parameters} summarizes the results from the gamma model 
from different data sets and shows the standard error of the year
effect increases for both the men's final and men's pooled data sets when
including 2022.  The men's final data venue effect error increased from 0.439
to 0.504 and from 0.0495 to 0.0554 for the men's pooled data.  This demonstrates
that the inclusion of 2022 increased the randomness of the data and motivates
the rank-based study (see Section~\ref{subsec:Results_Rank}).

@@

Using the gamma function with link set to log, the probability of the reaction 
time being less than 0.1 seconds can be calculated to see if Allen's 
disqualification was an anomaly as described in Section~\ref{sec:glmm}.
Calculations were performed in R using the methods described in 
section~\ref{sec:glmm} with the use of the \citet{lme4} package.

At a size of $n=10,000,000$, \eds{use different letter as $n$ was already used 
for number of athletes, $B?$} \of{Could I just say "At a size of 10,000,000"? 
Otherwise I am fine with using "B" instead of "n"} \eds{we may have lost the 
rest of this sentence?}

@@

%Citation needed in this section
Since 1985 Seiko Holding Corporations has served every World Athletics 
Championships as the official timer \citep{wa2022seiko}.  Since 1985, the 
technology and the ability
to accurately predict measurements: long jump distances, false starts, total 
time, reaction time, etc. have all dramatically improved.  Seiko did not start 
tracking reaction time as an official measurement until 1999 in the men's and 
women's 100 meter \eds{110 for men?} hurdles. 
Seiko regularly updates their equipment so that they provide cutting edge
technology to the World Track and Field Championships, the highest stakes in the 
world of running outside of the Olympics. Seiko's technology for detecting 
reaction times relies on the pressure that athletes exert on the plate when they 
push off.  Their systems measure the time differential
between when the ``gun" goes off to start the race and when the pressure changes 
\citep{wa2022seiko}.  If an athlete has a reaction time under 0.1 seconds, it 
is deemed a 
false start as it is considered that no athlete can react so quickly 
\citep{Seiko-Timing}.  Thus, a reaction time of for example 0.05, suggests that 
the athlete predicted the gun and it was luck that caused their abnormally high 
time.  At the World Championships level, World Athletics and Seiko want to remove 
that element of luck and thus impose the 0.1 second barrier. In 2013, Seiko 
upgraded their timing equipment for sprinting events (110 meter hurdles falls 
under this category) \citep{wa2013backtage}. Since 1999, the
slowest reaction 
times in the men's 110-meter hurdle were recorded in 2013.  That is not to say
 that the higher 
times were caused by Seiko's equipment, but the two may be related.  It is worth 
noting that prior
to 2022, Seiko again upgraded its technology but for its jump management system.
\eds{I would argue that this paragraph should go in the discussion as there are 
no results presented here}
